Quantum Interval-Valued Probability: Contextuality and the Born Rule

### 2018-03-31 revised_prop_letter_ajhs.tex
revised_prop_letter_ajhs.tex Fix suggestion from Andy...

\noindent The first three constraints,
Eqs.~(\ref{eq:QIVPM-constraints}), are the direct counterpart of the
corresponding ones for classical IVPMs.  {\color{red}Note that the minus sign
appears in Eq.~(\ref{eq:complement}) is accommodated by the $x\le0$ case in
Eq.~(\ref{eq:interval-times}).}  With the understanding that
the union of classical sets $E_1\cup E_2$ is replaced by
$P_0+P_1-P_0P_1$ in the case of quantum projection
operators~\cite{Griffiths2003}, the last condition,
Eq.~(\ref{eq:QuantumInterval-valuedProbability-Inclusion}), is a
direct counterpart of the convexity condition of
Eq.~(\ref{eq:classicalconvex}). Thus our definition of QIVPMs merges
aspects of both classical IVPMs and quantum probability measures.

@@

We next prove a main result of this paper which is a stronger
statement of contextuality that includes the effects of
finite-precision.  Every QIVPM must map some events to truly uncertain
intervals, not just ``almost definite intervals.'' The proof requires
two simple lemmas which we present first.

@@

\begin{tabular}{cc}
\toprule 
\addlinespace
$P$ & $\bar{\mu}\left(P\right)$\tabularnewline\addlinespace
\midrule
\midrule 
\addlinespace
$\mathbb{0}$, $\proj{0}$, $\proj{\ps}$, $\proj{\ps'}$ & $\imposs$\tabularnewline\addlinespace
\midrule 
\addlinespace
$\mathbb{1}$, $\mathbb{1}-\proj{0}$, $\mathbb{1}-\proj{\ps}$, 
$\mathbb{1}-\proj{\ps'}$ & $\necess$\tabularnewline\addlinespace
\midrule 
\addlinespace
All other projectors & $\unknown$\tabularnewline\addlinespace
\bottomrule
\end{tabular}
\end{table}
Next we will prove by contradiction that
$\coreBorn\left(\bar{\mu},\events\right)$ is the empty set. Suppose
there is a state
$\rho=\sum_{j=1}^{N}q_{j}\proj{\phi_{j}}\in\coreBorn\left(\bar{\mu},\events\right)$,
where $\sum_{j=1}^{N}q_{j}=1$ and $q_{j} > 0$. Since
we assumed the core $\coreBorn\left(\bar{\mu},\events\right)$ 
is non-empty, so  $\muB_{\rho}(P)\in\bar{\mu}(P)$, and 
Table~\ref{tab:non-Born-QIVPM} tells us that
$\bar{\mu}(\proj{0})=\imposs=[0,0]$, we must conclude that
$\muB_{\rho}(\proj{0})=0\in[0,0]$, and similarly for $\proj{\ps}$ and $\proj{\ps'}$.
If this is true, then $\ip{\phi_{j}}{0}=\ip{\phi_{j}}{\ps}=\ip{\phi_{j}}{\ps'}=0$
for all~$j$. {\color{red}Since any vector can be decomposed in terms of the
linearly independent basis $\left\{\ket{0},\ket{\ps},\ket{\ps'}\right\}$, all
$\ket{\phi_{j}}$ are perpendicular to any vector in the Hilbert space, and so
are not valid non-empty states, violating the assumption
$\ip{\phi_{j}}{\phi_{j}}=1$, and thus the theorem is proved.}
\yutsung{The above new text seems not very smooth. How about the following?}
\noindent If this is true, then $\ip{0}{\phi_{j}}=\ip{\ps}{\phi_{j}}=\ip{\ps'}{\phi_{j}}=0$
for all~$j$. This follows that
\begin{subequations}
\begin{eqnarray}
 &  & \ip{1}{\phi_{j}}=\sqrt{2}\ip{\ps}{\phi_{j}}-\ip{0}{\phi_{j}}=0\,,\\
 &  & \ip{2}{\phi_{j}}=\sqrt{2}\ip{\ps'}{\phi_{j}}-\ip{0}{\phi_{j}}=0\,.
\end{eqnarray}
\end{subequations}
The above equation implies
$\ket{\phi_{j}}=\ket{0}\ip{0}{\phi_{j}}+\ket{1}\ip{1}{\phi_{j}}+\ket{2}\ip{2}{\phi_{j}}=0$,
violating the assumption that $\ket{\phi_{j}}$ is a normalized state, and thus the
theorem is proved.
\yutsung{End of the suggested text.}

@@

To that end, we {\color{red}introduce the conept of \emph{interval maps} which
we can use to construct a consistent family of QIVPM.}
A{\color{red}n} interval map $f:\left[0,1\right]\rightarrow\mathscr{I}$
maps every real-valued probability $x\in\left[0,1\right]$ to a set
of intervals $f\left(x\right)=\left[\ell,r\right]$ containing $x$,
where $\left[0,1\right]$ denotes the set of real-valued probabilities
{\color{red}(this is unrelated to the interval-valued probability
$\unknown$).} \yutsung{How about change to ``(this should not be confused with
the interval-valued probability $\unknown$)''?}
We also need a notion of {\color{red}\emph{norm}} to quantify the
distance between (pure or mixed) states. The norm of a pure state
$\rho=\proj{\psi}$ is defined as usual by $\left\Vert \psi\right\Vert =\sqrt{\ip{\psi}{\psi}}$.
{\color{red} For any given Hermitian operator~$A$, we choose}
operator norm $\left\Vert A\right\Vert =\max_{\left\Vert \psi\right\Vert =1}\left\Vert A\ket{\psi}\right\Vert $,
which is also known as the $2$-norm or the spectral norm~\cite{RobertsVarberg1973,peres1995quantum,GolubVanLoan1996,Foucart2012}.
In fact, for any such matrix, including the density matrix~$\rho$,
th{\color{red}is norm is the eigenvalue with maximum absolute value}. Then, a
finite-precision extension of Gleason's theorem can be stated as
follows{\color{red}:}

@@

A{\color{red}n} interval map is called \emph{ultramodular} if it satisfies
the following properties.

@@

Statement~\ref{enu:iota-subject-to} implies \ref{enu:iota-mu-CIVPM}
and~\ref{enu:iota-mu-QIVPM} as we have outlined above. Conversely, for the quantum case, we want
to show that if $\ultramodular$ is not ultramodular, then for
some quantum probability measure $\mu$, the composite
$\ultramodular\circ\mu$ might not be a QIVPM. Suppose there are three
particular numbers~$x_{0}$, $x_{1}$, and $x_{2}\in\left[0,1\right]$
such that $y=x_{0}+x_{1}+x_{2}\in\left[0,1\right]$, but they don't
satisfy Eq.~(\ref{eq:iota-Inclusion}). Consider the state:
\[
\rho=x_{0}\proj{0}+x_{1}\proj{1}+x_{2}\proj{2}+\left(1-y\right)\proj{3}\,.
\]
The induced map using the Born rule and blurred\yutsung{A ``blurred'' here.} by $\ultramodular$,
$\ultramodular\circ\muB_{\rho}$, fails to satisfy
Eq.~(\ref{eq:QuantumInterval-valuedProbability-Inclusion}) when
$P_{0}=\proj{0}+\proj{2}$ and $P_{1}=\proj{1}+\proj{2}$. In other
words, this induced map fails to be a QIVPM. For the classical case, if
$\ultramodular$ is not ultramodular, we also want to find a classical
probability measure $\mu:\eventsC\rightarrow\left[0,1\right]$ such
that $\ultramodular\circ\mu$ is not a classical IVPM. This can be
done by restricting our previous quantum probability measure
$\muB_{\rho}$ to the space of events  $\eventsC$  generated by the
mutually commuting projectors $\proj{0}$, $\proj{1}$, $\proj{2}$, and
$\proj{3}$.  The restricted function
$\mu=\muB_{\rho}|_{\eventsC}$ is then a classical probability
measure, and the induced map $\ultramodular\circ\mu$ fails to be
a classical IVPM for the same reason as in the quantum case.

In other words, essential properties of QIVPMs constructed using
interval maps can be gleaned from the properties of
ultramodular functions. The following is a most\yutsung{``a most''?} interesting property
in our setting{\color{red}:}

@@

\andy{no ending? What was the point of the section?}
\yutsung{Begin of comments}
Thm.~\ref{thm:convex-uncountable} proved
$\ultramodularL:\left[0,1\right]\rightarrow\left[0,1\right]$ is continuous, and
$\mathscr{I}$ contains uncountably many intervals. In the sense of ``achieving
computable theory of quantum mechanics'', this fact may relate the following
Thm.~\ref{thm:rho2rho-continuous} \cite{Ziegler2007,weihrauch2012computable}
whose idea is that when we mapping between continuums, only continuous functions
could be (computably-)approximated, while discontinuous functions might be too
wild to be approximated.

@@

\begin{thm}\label{thm:rho2rho-continuous}
If $f:\left[0,1\right]\rightarrow\mathbb{R}$ is
($\rho\rightarrow\rho$)-computable, then $f$ is continuous.
\end{thm}
\yutsung{End of comments}

@@

% \input{revised_prop_letter_ajhs.bbl}
\bibliography{prop}
\end{document}

>>>>

\noindent The first three constraints,
Eqs.~(\ref{eq:QIVPM-constraints}), are the direct counterpart of the
corresponding ones for classical IVPMs.  {\color{red}Note that the minus sign
appearing in Eq.~(\ref{eq:complement}) is accommodated by the $x\le0$ case in
Eq.~(\ref{eq:interval-times}).}  With the understanding that
the union of classical sets $E_1\cup E_2$ is replaced by
$P_0+P_1-P_0P_1$ in the case of quantum projection
operators~\cite{Griffiths2003}, the last condition,
Eq.~(\ref{eq:QuantumInterval-valuedProbability-Inclusion}), is a
direct counterpart of the convexity condition of
Eq.~(\ref{eq:classicalconvex}). Thus our definition of QIVPMs merges
aspects of both classical IVPMs and quantum probability measures.

@@

We next prove a main result of this paper which is a stronger
statement of contextuality that includes the effects of
finite-precision.  Every QIVPM must map some events to truly uncertain
intervals, not just ``almost definite intervals.'' The proof requires
two simple lemmas {\color{red}that} we present first.

@@

\begin{tabular}{cc}
\toprule 
\addlinespace
$P$ & $\bar{\mu}\left(P\right)$\tabularnewline\addlinespace
\midrule
\midrule 
\addlinespace
$\mathbb{0}$, $\proj{0}$, $\proj{\ps}$, $\proj{\ps'}$ & $\imposs$\tabularnewline\addlinespace
\midrule 
\addlinespace
$\mathbb{1}$, $\mathbb{1}-\proj{0}$, $\mathbb{1}-\proj{\ps}$, 
$\mathbb{1}-\proj{\ps'}$ & $\necess$\tabularnewline\addlinespace
\midrule 
\addlinespace
All other projectors & $\unknown$\tabularnewline\addlinespace
\bottomrule
\end{tabular}
\end{table}
Next we will prove by contradiction that
$\coreBorn\left(\bar{\mu},\events\right)$ is the empty set. Suppose
there is a state
$\rho=\sum_{j=1}^{N}q_{j}\proj{\phi_{j}}\in\coreBorn\left(\bar{\mu},\events\right)$,
where $\sum_{j=1}^{N}q_{j}=1$ and $q_{j} > 0$. Since
we assumed the core $\coreBorn\left(\bar{\mu},\events\right)$ 
is non-empty, so  $\muB_{\rho}(P)\in\bar{\mu}(P)$, and 
Table~\ref{tab:non-Born-QIVPM} tells us that
$\bar{\mu}(\proj{0})=\imposs=[0,0]$, we must conclude that
$\muB_{\rho}(\proj{0})=0\in[0,0]$, and similarly for $\proj{\ps}$ and $\proj{\ps'}$.
If this is true, then {\color{red}$\ip{0}{\phi_{j}}=\ip{\ps}{\phi_{j}}=\ip{\ps'}{\phi_{j}}=0$
for all~$j$, and thus
\begin{subequations}
\begin{eqnarray}
 &  & \ip{1}{\phi_{j}}=\sqrt{2}\ip{\ps}{\phi_{j}}-\ip{0}{\phi_{j}}=0\,,\\
 &  & \ip{2}{\phi_{j}}=\sqrt{2}\ip{\ps'}{\phi_{j}}-\ip{0}{\phi_{j}}=0\,.
\end{eqnarray}
\end{subequations}
The above equations imply
$\ket{\phi_{j}}=\ket{0}\ip{0}{\phi_{j}}+\ket{1}\ip{1}{\phi_{j}}+\ket{2}\ip{2}{\phi_{j}}=0$,
violating the assumption that $\ket{\phi_{j}}$ is a normalized state, and thus the
theorem is proved.}

@@

To that end, we {\color{red}introduce the concept of \emph{interval maps}, which
we can use to construct a consistent family of QIVPMs.}
A{\color{red}n} interval map $f:\left[0,1\right]\rightarrow\mathscr{I}$
maps every real-valued probability $x\in\left[0,1\right]$ to a set
of intervals $f\left(x\right)=\left[\ell,r\right]$ containing $x$,
where $\left[0,1\right]$ denotes the set of real-valued probabilities
{\color{red}(this should not be confused with
the interval-valued probability $\unknown$)}.
We also need a notion of {\color{red}\emph{norm}} to quantify the
distance between (pure or mixed) states. The norm of a pure state
$\rho=\proj{\psi}$ is defined as usual by $\left\Vert \psi\right\Vert =\sqrt{\ip{\psi}{\psi}}$.
{\color{red} For any given Hermitian operator~$A$, we choose the}
operator norm $\left\Vert A\right\Vert =\max_{\left\Vert \psi\right\Vert =1}\left\Vert A\ket{\psi}\right\Vert $,
which is also known as the $2$-norm or the spectral norm~\cite{RobertsVarberg1973,peres1995quantum,GolubVanLoan1996,Foucart2012}.
In fact, for any such matrix, including the density matrix~$\rho$,
th{\color{red}is norm is the eigenvalue with maximum absolute value}. Then, a
finite-precision extension of Gleason's theorem can be stated as
follows{\color{red}:}

@@

A{\color{red}n} interval map is called \emph{ultramodular} if it satisfies
the following properties{\color{red}:}

@@

Statement~\ref{enu:iota-subject-to} implies \ref{enu:iota-mu-CIVPM}
and~\ref{enu:iota-mu-QIVPM} as we have outlined above. Conversely, for the quantum case, we want
to show that if $\ultramodular$ is not ultramodular, then for
some quantum probability measure $\mu$, the composite
$\ultramodular\circ\mu$ might not be a QIVPM. Suppose there are three
particular numbers~$x_{0}$, $x_{1}$, and $x_{2}\in\left[0,1\right]$
such that $y=x_{0}+x_{1}+x_{2}\in\left[0,1\right]$, but they don't
satisfy Eq.~(\ref{eq:iota-Inclusion}). Consider the state:
\[
\rho=x_{0}\proj{0}+x_{1}\proj{1}+x_{2}\proj{2}+\left(1-y\right)\proj{3}\,.
\]
The induced map {\color{red}$\ultramodular\circ\muB_{\rho}$ constructed using the Born rule and
blurred by $\ultramodular$ fails} to satisfy
Eq.~(\ref{eq:QuantumInterval-valuedProbability-Inclusion}) when
$P_{0}=\proj{0}+\proj{2}$ and $P_{1}=\proj{1}+\proj{2}$. In other
words, this induced map fails to be a QIVPM. For the classical case, if
$\ultramodular$ is not ultramodular, we also want to find a classical
probability measure $\mu:\eventsC\rightarrow\left[0,1\right]$ such
that $\ultramodular\circ\mu$ is not a classical IVPM. This can be
done by restricting our previous quantum probability measure
$\muB_{\rho}$ to the space of events  $\eventsC$  generated by the
mutually commuting projectors $\proj{0}$, $\proj{1}$, $\proj{2}$, and
$\proj{3}$.  The restricted function
$\mu=\muB_{\rho}|_{\eventsC}$ is then a classical probability
measure, and the induced map $\ultramodular\circ\mu$ fails to be
a classical IVPM for the same reason as in the quantum case.

In other words, essential properties of QIVPMs constructed using
interval maps can be gleaned from the properties of
ultramodular functions. The following is a most interesting property
in our setting{\color{red}:}

@@

\begin{comment}
\andy{no ending? What was the point of the section?}
\yutsung{Begin of comments}
Thm.~\ref{thm:convex-uncountable} proved
$\ultramodularL:\left[0,1\right]\rightarrow\left[0,1\right]$ is continuous, and
$\mathscr{I}$ contains uncountably many intervals. In the sense of ``achieving
computable theory of quantum mechanics'', this fact may relate the following
Thm.~\ref{thm:rho2rho-continuous} \cite{Ziegler2007,weihrauch2012computable}
whose idea is that when we mapping between continuums, only continuous functions
could be (computably-)approximated, while discontinuous functions might be too
wild to be approximated.

@@

\begin{thm}\label{thm:rho2rho-continuous}
If $f:\left[0,1\right]\rightarrow\mathbb{R}$ is
($\rho\rightarrow\rho$)-computable, then $f$ is continuous.
\end{thm}
\yutsung{End of comments}
\end{comment}

@@

\input{revised_prop_letter_ajhs.bbl}
\begin{comment}
\bibliography{prop}
\end{comment}
\end{document}
